\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

The Internet forwards packets without regard for why they were sent, so the network optimizes a target
that rarely matches what an application actually values. Path-aware architectures such as SCION change
this by letting endpoints choose among multiple end-to-end paths, placing the path decision in the hands
of the one party that knows the application's intent. This thesis studies how to make that choice well:
how to select the best path -- or the best combination of paths -- given both the application's intent and
the current state of the network.

We take a single idea from recent congestion-control work, that application intent should be a first-class
input which reshapes a network mechanism, and push it into two settings. First, we turn a fixed-objective
DQN path selector for SCION into an intent-conditioned one, combining a permutation-equivariant scoring
architecture that handles a variable, changing set of candidate paths with feature-wise modulation of path
features by an intent vector; a single policy then serves a spectrum of objectives -- throughput,
latency, trust, probe-frugality -- without retraining. We show, analytically and empirically, that
conditioning must \emph{modulate} the decision rather than merely accompany it, since an additive signal
cancels under the comparison that selects a path. Second, we ask what happens when several paths are used
at once and the application is optimized alongside the network: a two-timescale hierarchical pair of
reinforcement-learning agents co-optimizes a real-time video call's encoding bitrate and its per-path
traffic split over Multipath QUIC. Here the central finding is that the value of learned multipath
scheduling depends sharply on network volatility -- nearly negligible when paths are stable, but decisive
under path churn and congestion bursts, where controlling the application alone collapses. Together the two
parts argue that on a path-aware Internet, application intent should steer every controllable decision
between the application and the wire, and they chart what remains to make such control deployable.
