\chapter{Conclusion}
\label{ch:conclusion}

This thesis asked how an endpoint on a path-aware Internet should choose the best use of its paths given
the intention of the application and the state of the network. It answered in two movements that share one
idea, drawn from Mortise~\cite{Shen2026}: treat application intent as a first-class input that reshapes a
network mechanism, not as a constant compiled into it.

Part~I turned a fixed-objective DQN path selector for SCION~\cite{Bourak2025} into an intent-conditioned
one. Two design decisions carried the work: a permutation-equivariant \emph{scoring} architecture that
handles a variable, changing set of candidate paths, and a \emph{feature-wise modulation} of path features
by the intent vector that lets a single policy serve the whole spectrum of objectives without retraining.
Part~II asked what happens if we use several paths at once and optimize the application itself, and built a
two-timescale hierarchical agent pair that co-optimizes video bitrate and per-path traffic split over
Multipath QUIC, coupled so that each agent accounts for the other.

\paragraph{What we learned.} The most useful findings were the ones that cut against intuition. First, that
\emph{how} intent is injected matters more than \emph{that} it is injected: for a comparison-based selector,
appending the intent to the state leaves decisions unchanged, because a symmetric signal cancels under the
argmax; only modulation reorders the alternatives. Second, that the value of learned multipath scheduling is
\emph{not} a fixed quantity but a function of network volatility -- negligible when paths are stable, where a
good bitrate controller over a trivial split suffices, yet decisive under churn and bursts, where the same
app-only configuration collapses. Both findings sharpen when, and when not, to reach for the more
sophisticated mechanism.

\paragraph{Future work.} Several threads remain open, and much of the draft's structure is deliberately laid
out to guide them. On the experimental side: completing the Part~I baseline, intent-alignment, and ablation
tables; adding seed-level statistics and the flat-versus-scoring comparison in Part~II; and porting the
non-stationary dynamics into the NS-3 backend so packet-level results can confirm the mock findings. On the
system side: replacing the TCP-like subflows with a real Multipath QUIC transport, and enriching the VMAF
surrogate so quality responds to delivery conditions, not just bitrate. On the research side, three
directions stand out: \emph{inferring} intent from traffic rather than assuming it is declared; \emph{online
and safe} learning against live traffic with bounded worst-case behavior; and pushing the ``one principle,
every controllable decision'' recipe further -- for instance, conditioning the Part~II agents on an explicit
intent vector as Part~I does, unifying the two systems into a single intent-conditioned, cross-layer
controller for path-aware networks.
