\chapter{Introduction}
\label{ch:introduction}

The Internet was designed to deliver packets, not experiences. A router forwards a datagram
toward its destination without any notion of \emph{why} the datagram was sent: whether it
carries a frame of an interactive video call that will be worthless if it arrives late, a
byte of a background software update that only cares about aggregate throughput, or a
control message whose loss would be catastrophic. For decades this deliberate ignorance --
the end-to-end principle -- was a feature. Yet as applications have become more diverse and
more demanding, the mismatch between what the network optimizes and what the application
actually values has grown into one of the central inefficiencies of modern networking.

Path-aware Internet architectures such as SCION~\cite{Perrig2017} change the terms of this
problem. In SCION, endpoints discover \emph{multiple} end-to-end paths to a destination and
choose among them, rather than deferring blindly to whatever route BGP happens to install.
This turns path selection from a network-internal decision into an application-facing one:
for the first time the entity that knows the application's intent -- the endpoint -- also
holds the lever that determines how traffic is carried. The question this thesis studies is
the natural consequence of that shift:

\begin{quote}
\itshape How should an endpoint choose the best path -- or the best \emph{combination} of
paths -- given both the intention of the application and the current state of the network?
\end{quote}

Answering this question is hard for three compounding reasons. First, \emph{``best'' is not
a fixed target}. A path that is best for a bulk transfer (high goodput, tolerant of latency)
is often a poor choice for a video call (which prizes low latency and jitter and will trade
raw throughput to get them). Any single hand-tuned objective is therefore wrong for some
application, and training a separate policy for every conceivable objective does not scale.
Second, \emph{the decision space is dynamic and unbounded}. The set of available paths between
two hosts differs from pair to pair and changes over time as links appear, fail, congest, or
recover; a policy that assumes a fixed, enumerated action set cannot cope. Third, \emph{the
state is only partially and expensively observable}. Fresh per-path measurements cost probing
overhead, so a good policy must decide well while measuring little.

This thesis attacks the question in two movements, unified by a single idea borrowed from
recent work on congestion control. The Mortise system~\cite{Shen2026} shows that a transport
mechanism can be steered to maximize an application's quality of experience (QoE) by treating
the application's \emph{intent} as an explicit input that reshapes how the mechanism is tuned,
rather than as a fixed constant baked into the algorithm. We take the same stance and push it
into two new settings.

\paragraph{Part~I: steering \emph{which} path.} We build on the DQN-based SCION path selector
of Bourak~et~al.~\cite{Bourak2025} and rework it into an intent-conditioned path-selection
agent. We improve the underlying SCION simulation environment, replace the fixed-action DQN
with a permutation-equivariant \emph{path-scoring} architecture that handles a variable number
of candidate paths, and -- the key contribution -- condition the agent on a compact
representation of application intent using \emph{feature-wise linear modulation}
(FiLM)~\cite{Perez2018}. A single trained policy can then serve a throughput-hungry transfer,
a latency-sensitive call, or any point in between, simply by changing the intent vector at
inference time, without retraining. This is the path-selection analogue of Mortise's
intent-driven parameter tuning.

\paragraph{Part~II: steering \emph{how much} of each path, and \emph{what} the application
sends.} Part~I chooses a single path. But SCION's multipath nature invites a more ambitious
question: what if we use several paths at once, and co-optimize the application itself? Where
Mortise tunes the congestion-control algorithm to the application, we tune both the
\emph{application} (the video encoder's target bitrate) and the \emph{network} (the split of
traffic across paths) jointly. Using a Multipath-QUIC~\cite{DeConinck2017} transport carrying
a real-time WebRTC-style video flow, we design a hierarchical pair of reinforcement-learning
agents -- a slow application agent that adapts bitrate and a fast transport agent that
schedules bytes across paths -- coupled so that each accounts for the other. This closes the
loop from intent all the way to what the application chooses to produce. Because Part~II's
Multipath~QUIC transport rides on top of the same path-aware substrate, it is best read as
\emph{multipath transport enabled by SCION's native path multiplicity}: SCION supplies the
several disjoint paths, and the agents decide how to use them.

Throughout, a set of questions guides the reader. \emph{Can one policy serve many intents
without retraining, and does conditioning actually change its decisions or merely its
rhetoric?} \emph{When is it worth splitting traffic across paths at all, and when does a
single good path suffice?} \emph{Does co-optimizing the application with the network beat
optimizing either alone?} \emph{And how far is any of this from something one could deploy?}
We will see that the answers are often not the obvious ones -- for instance, that multipath
scheduling can be nearly worthless in a stable network yet decisive in a volatile one.

\section{Contributions}
\label{sec:intro:contributions}

This thesis makes the following contributions.

\begin{enumerate}
  \item \textbf{An intent-conditioned path-selection agent for SCION.} We extend the DQN path
  selector of Bourak~et~al.~\cite{Bourak2025} with (i) an improved, higher-fidelity SCION
  simulation environment driven by realistic beaconing and multi-day traffic, (ii) a
  permutation-equivariant path-scoring network that supports a variable and changing set of
  candidate paths, and (iii) a FiLM-based conditioning mechanism that lets one policy realize a
  spectrum of application intents without retraining (\cref{ch:p1design,ch:p1impl}).

  \item \textbf{Evidence that intent conditioning must modulate, not merely append.} We show
  analytically and empirically why naively concatenating an intent vector to the state fails to
  change a greedy selector's decisions, and why feature-wise modulation is the appropriate
  mechanism (\cref{ch:p1design}).

  \item \textbf{A hierarchical agent pair for joint application--network optimization over
  multipath.} We design and implement a two-timescale reinforcement-learning system that
  co-optimizes video bitrate (application) and per-path traffic split (network) over
  Multipath~QUIC, realized on both a packet-level NS-3 backend and a fast analytical backend
  behind a common interface (\cref{ch:p2design,ch:p2impl}).

  \item \textbf{An evaluation and analysis of \emph{when} each mechanism pays off.} We
  characterize the conditions under which intent conditioning and multipath scheduling deliver
  value -- and the conditions under which simpler baselines suffice -- and we assess the
  overhead, scalability, generalization, and deployability of both systems
  (\cref{ch:analysis,ch:discussion}).
\end{enumerate}

\section{Organization}
\label{sec:intro:organization}

\Cref{ch:background} reviews the background this thesis builds on: the SCION architecture, the
reinforcement-learning machinery (DQN and its refinements, soft actor-critic, variable
action spaces, and FiLM-based conditioning), and the QoE and multipath-transport concepts used
in Part~II. \Cref{ch:problem} states the problem precisely, lists the properties a good
solution must have, and makes explicit the assumptions and trust model.

The body is then organized in three parts. \textbf{Part~I} covers intent-aware single-path
selection in SCION: its design (\cref{ch:p1design}), implementation (\cref{ch:p1impl}), and
evaluation (\cref{ch:p1eval}). \textbf{Part~II} covers joint application--network optimization
over multipath: design (\cref{ch:p2design}), implementation (\cref{ch:p2impl}), and evaluation
(\cref{ch:p2eval}). \textbf{Part~III} draws the two together: \cref{ch:analysis} analyzes
overhead, scalability, generalization, and deployability; \cref{ch:discussion} discusses what
the results mean; \cref{ch:related} places the work in the research landscape; and
\cref{ch:conclusion} concludes.
