\chapter{Discussion and System Synthesis}
\label{ch:synthesis}

This chapter steps back from the two systems to their joint meaning. \Cref{ch:discussion} synthesizes
what intent-conditioned single-path selection and joint multipath co-optimization say together about
intent-driven network control. \Cref{ch:analysis} then analyzes both systems along the axes that decide
whether learned path control is practical -- overhead, scalability, and generalization -- and
\cref{sec:analysis:deployability} states the deployment gaps and trust assumptions under which the
controllers operate. Where a claim depends on measurements still in progress, it is marked.

\section{Synthesis of Findings}
\label{ch:discussion}

Read together, the two systems tell a consistent story about \emph{when} additional control machinery
earns its keep.

\subsection{Interpretation}
\label{sec:disc:interpretation}

\paragraph{Intent conditioning turns a family of policies into one.} The central lesson of
\cref{ch:single_path} is that the right way to serve many objectives is not many models but one model that
takes the objective as input -- provided the objective is allowed to \emph{modulate} the decision rather
than merely accompany it. The concatenation-versus-FiLM contrast (\cref{sec:p1design:film}) is the crisp
version of this point: a conditioning signal that enters symmetrically across the alternatives cancels
under the comparison that picks the action, so the policy can look conditioned while behaving identically.
Feature-wise modulation avoids this because it changes the \emph{relative} salience of path features, and
thus which path wins. This is a design principle, not a SCION-specific trick: any comparison-based selector
conditioned on an objective should modulate its per-alternative features.

\paragraph{The value of multipath control is a function of volatility.} The central lesson of
\cref{ch:multipath} is that learned multipath scheduling ranges from nearly worthless to decisive
depending on the network. In the static scenario, a good bitrate controller over a trivial split captured
essentially all achievable QoE, and the learned transport agent added little (\cref{sec:p2eval:results}).
In the dynamic scenario, the same app-only configuration collapsed to negative QoE while the learned split
sustained a usable call. The practical implication is that one should not evaluate multipath schedulers
only in stable conditions -- that is precisely the regime in which they look unnecessary -- and that the
case for learned scheduling rests on volatility, churn, and bursts, where static policies cannot adapt.

\paragraph{One principle, two layers.} Both systems instantiate the Mortise principle~\cite{Shen2026} --
intent as a first-class input that reshapes a network mechanism -- at successively deeper points of
control. The single-path selector reshapes \emph{which path} is chosen; the multipath system reshapes both
\emph{how the paths are used} and \emph{what the application produces}. The progression suggests a general
recipe: expose the application's intent, and let it steer every controllable decision between the
application and the wire.

\subsection{Implications}
\label{sec:disc:implications}

The results matter for how we build controllers on path-aware architectures. Prior learned path
selectors~\cite{Bourak2025} and learned congestion or scheduling controllers~\cite{Jay2019, Zhang2019,
Wu2020} each optimize a fixed objective; our work argues that on a path-aware substrate the objective should
be a knob, and that a single conditioned policy can replace a zoo of per-objective models. For real-time
media specifically, the joint bitrate-and-split result reinforces a theme from learned
ABR~\cite{Mao2017} -- that application and network decisions are better made together -- and extends it to
the multipath setting that SCION makes routine. More broadly, the two systems are evidence that SCION's
endpoint path choice is not merely a routing convenience but an enabler of application-aware optimization
that is awkward or impossible in a destination-based Internet.

\subsection{Limitations}
\label{sec:disc:limitations}

The results are bounded in ways that the analysis (\cref{ch:analysis}) details and that fairness requires
restating here. Both systems are evaluated in simulation, and the multipath system's headline
dynamic-scenario numbers are from the mock backend pending NS-3 confirmation (\cref{sec:p2impl:gaps}).
Intent in this thesis is \emph{declared}, not \emph{inferred}: we assume the application can state its
objective, and we do not study how to recover a latent intent from traffic -- a natural but separate
problem. The transport model is TCP-like rather than true Multipath QUIC, and the VMAF surrogate is
bitrate-dominated, so absolute QoE figures should be read as relative comparisons rather than calibrated
user scores. Finally, we have not addressed online, safe learning against live traffic, which any
deployment would require. None of these undercut the two qualitative findings -- modulation-based
conditioning works where concatenation fails, and multipath control pays off in proportion to volatility --
but they do delimit how far the quantitative results should be pushed.

\section{System Analysis}
\label{ch:analysis}

This section analyzes the two systems along the dimensions that determine whether learned path control is
practical: overhead (\cref{sec:analysis:overhead}), scalability (\cref{sec:analysis:scalability}), and
generalization (\cref{sec:analysis:generalization}).

\subsection{Overhead}
\label{sec:analysis:overhead}

Two kinds of overhead matter: measurement and computation.

\paragraph{Measurement (probing) overhead.} In the single-path selector (\cref{ch:single_path}), probing
is a first-class cost charged in the reward (\cref{eq:p1reward}) and accounted per decision
(\cref{sec:p1impl:baselines}). The learned selector probes only the paths it inspects, in contrast to
heuristics that require a metric on every path, so its measurement footprint is a fraction of an exhaustive
scan at comparable selection quality -- the property confirmed in \cref{sec:p1eval:overhead}. In the
multipath system (\cref{ch:multipath}), per-path state is derived from ordinary transport feedback (RTT
samples, acknowledgements, loss signals) rather than dedicated probes, so the marginal measurement cost of
the split decision is low.

\paragraph{Computational overhead.} Both policies are small neural networks evaluated once per decision.
The single-path selector runs at path-selection granularity (seconds to minutes) and is comfortably cheap.
The multipath transport agent is the tighter constraint: it runs \emph{per frame}
(${\approx}\SI{33}{\milli\second}$ at 30\,fps), so its inference must fit well inside a frame interval. The
scoring architecture keeps this affordable -- a shared per-path encoder evaluated $N$ times and a softmax --
but the per-frame budget is the binding real-time constraint. \gap{report measured per-decision inference
latency for both agents, and confirm the transport agent's inference fits within the frame budget on the
target hardware.}

\subsection{Scalability}
\label{sec:analysis:scalability}

The scoring architecture is what makes both systems scale in the number of paths. Because a shared network
scores each path independently and decisions are made by comparison (\cref{sec:p1design:scoring}) or softmax
(\cref{sec:p2design:scoring}), the parameter count is independent of $N$, and inference is linear in $N$.
This lets one trained model serve source--destination pairs with any number of paths and topologies of any
size, without retraining or padding to a fixed maximum -- the limitation of the flat agents. The remaining
scalability question is training coverage: the environment must expose enough variety of path counts,
topologies, and conditions for the shared scorer to generalize. \gap{report how selection quality holds as
topology size and path count grow, ideally on topologies larger than those seen in training.}

\subsection{Generalization}
\label{sec:analysis:generalization}

Generalization is required along three axes. \textbf{Across intents:} the FiLM conditioning of
\cref{ch:single_path} is meant to interpolate to intents between (and, ideally, beyond) the trained reward
profiles, so that a novel weight vector yields sensible behavior without retraining -- tested in
\cref{sec:p1eval:intent}. \textbf{Across topologies and path counts:} the permutation-equivariant design
gives structural generalization, but only empirical results can confirm it holds off the training
distribution. \textbf{Across conditions:} the dynamic scenario of \cref{ch:multipath}
(\cref{sec:p2eval:results}) is itself a generalization stress test, since the policy must cope with churn
and bursts rather than a stationary environment; the collapse of the fixed-schedule baselines there shows
that static policies do \emph{not} generalize across volatility, which is the gap learned control fills.
\gap{quantify off-distribution degradation for both systems -- unseen intents, larger topologies, and
unseen dynamics regimes -- to bound how gracefully performance decays.}

\section{Deployment Considerations and Trust Assumptions}
\label{sec:analysis:deployability}

\subsection{Deployment Gaps}
Several gaps still separate these prototypes from a deployment.

\begin{itemize}
  \item \textbf{From simulation to reality.} Both systems are trained in simulation. The single-path study
  relies on a synthetic traffic and beaconing environment; the multipath system on NS-3 and a mock
  transport. Bridging the gap to deployment would require sim-to-real transfer -- for example through domain
  randomization or online fine-tuning against live measurements -- and the packet-level NS-3 backend is the
  intended stepping stone (\cref{sec:p2impl:gaps}).
  \item \textbf{Real transport.} The multipath system models subflows with TCP-like congestion control
  rather than true Multipath QUIC~\cite{DeConinck2017}. A real QUIC integration is therefore needed to
  validate the per-frame scheduling under an actual multipath transport.
  \item \textbf{Online and safe learning.} A deployed agent would learn or adapt against live traffic, where
  exploration has real cost. Bounding worst-case behavior -- for example by falling back to a safe heuristic
  when the policy is uncertain, or by constraining exploration -- is a prerequisite that this work does not
  address.
\end{itemize}

Two properties help on the positive side. First, the endpoint-driven nature of SCION path selection means
these controllers require no network-internal changes. Second, the horizon-agnostic application agent
(\cref{sec:p2design:hierarchy}) is designed for unbounded real sessions rather than fixed-length episodes.

\subsection{Trust Assumptions}

This thesis is a performance and control study, not a security study, so in place of a classical
attacker model we state the trust assumptions under which the learned controllers operate and note
where security considerations would enter a real deployment.

\begin{itemize}
  \item \textbf{Honest measurements.} We assume that the per-path measurements an agent acts on --
  whether obtained by active probing (\cref{ch:single_path}) or transport feedback (\cref{ch:multipath}) --
  are honest. A malicious on-path element that fabricates favorable metrics could steer the agent toward a
  bad path; defending against that is out of scope, although the trust term below offers partial mitigation.
  \item \textbf{SCION path authenticity and trust.} SCION's cryptographically authenticated paths mean
  an agent knows \emph{which} ASes a path traverses and cannot be silently redirected. The single-path
  reward includes a \emph{trust} component derived from path quality, allowing the agent to prefer
  higher-confidence paths and discount those with anomalous loss or latency -- a lightweight hedge against
  unreliable or adversarial path advertisement, not a full defense.
  \item \textbf{No adversarial control of the learner.} We assume the agent's own training and
  inference pipeline is trusted; adversarial manipulation of the RL policy (e.g.\ reward poisoning or
  observation attacks) is beyond this work.
\end{itemize}

A production deployment would need to revisit these assumptions -- in particular by verifying measurement
integrity and by bounding the damage a single misreported path can do.
