\chapter{Related Work}
\label{ch:related}

Where \cref{ch:background} covered the systems this thesis \emph{builds on}, this chapter surveys work that
attacks the \emph{same or similar problems} and positions our contributions against it. We group it into
learned routing and traffic engineering (\cref{sec:rel:routing}), learned congestion control and multipath
scheduling (\cref{sec:rel:transport}), adaptive bitrate and QoE-driven control (\cref{sec:rel:abr}),
goal-conditioned and modulated policies (\cref{sec:rel:gcrl}), and intent-based networking
(\cref{sec:rel:ibn}).

\section{Learned Routing and Traffic Engineering}
\label{sec:rel:routing}

A line of work applies reinforcement learning to routing and traffic engineering. \emph{Learning to
Route}~\cite{Valadarsky2017} studies RL and learned representations for routing, and experience-driven
networking~\cite{Xu2018} uses deep RL for traffic-engineering decisions. These target network-operator
objectives (e.g.\ minimizing congestion globally) from a control-plane vantage point. Our single-path selector differs in
both the actor and the objective: the decision is made by an \emph{endpoint} choosing among
cryptographically authenticated SCION paths, and the objective is the \emph{application's} intent rather than
a global network utility. Most directly related is the DQN SCION path selector of
Bourak~et~al.~\cite{Bourak2025}, which we extend; the differences -- a variable-action scoring architecture
and FiLM-based intent conditioning that removes the single-objective limitation -- are the substance of \cref{ch:single_path}.

\section{Learned Congestion Control and Multipath Scheduling}
\label{sec:rel:transport}

Learned transport controllers include Aurora~\cite{Jay2019}, which casts congestion control as RL, and
Orca~\cite{Abbasloo2020}, which combines a learned layer with a classic controller. Mortise~\cite{Shen2026}
is the closest in spirit: it auto-tunes congestion-control parameters to the running application's QoE, and
its principle -- intent as an input that reshapes a mechanism -- is the template we transplant into path
selection and joint multipath control. For multipath \emph{scheduling}, ReLeS~\cite{Zhang2019} learns a
Multipath QUIC scheduler and Peekaboo~\cite{Wu2020} learns to schedule across heterogeneous paths in dynamic
environments; both optimize a scheduling policy for a given, exogenous load. The multipath system's distinction is that it
does \emph{not} treat the offered load as fixed: it co-optimizes the application's bitrate with the split, so
the controller can relieve congestion it would otherwise create -- a coupling those schedulers, and the
Multipath TCP mechanisms of RFC~8684~\cite{RFC8684}, do not address.

\section{Adaptive Bitrate and QoE-Driven Control}
\label{sec:rel:abr}

Learned adaptive bitrate control, exemplified by Pensieve~\cite{Mao2017}, showed RL beating hand-crafted ABR
heuristics for buffered video streaming, using QoE-shaped rewards and quality metrics such as
VMAF~\cite{VMAF2016}. Our multipath system shares the QoE-driven, learned-bitrate stance but targets \emph{real-time}
video (no playback buffer, deadline-based loss) and, crucially, couples the bitrate decision to a
simultaneous multipath scheduling decision. Where Pensieve adapts rate to a single, exogenous network path,
we adapt rate \emph{and} decide how to spread it across several paths, jointly.


\subsection{Mortise: intent-driven tuning of a transport mechanism}
The conceptual template for both parts is Mortise~\cite{Shen2026}. Mortise observes that a single
congestion-control algorithm cannot be optimal for every application, and instead \emph{auto-tunes}
the algorithm's parameters to maximize the running application's QoE, using the application's revealed
preferences over throughput, latency, and loss to steer the tuning. The essential idea -- treat
application intent as a first-class input that reshapes a network mechanism rather than a constant
compiled into it -- is exactly what we transplant. The single-path selector applies it to \emph{path selection} (which
path to use), and the multipath system applies it to \emph{joint rate and multipath scheduling} (what
the application sends and how it is spread across paths).

\section{Goal-Conditioned and Modulated Policies}
\label{sec:rel:gcrl}

The machinery for serving many objectives with one policy comes from goal-conditioned RL -- universal value
function approximators~\cite{Schaul2015} and hindsight relabeling~\cite{Andrychowicz2017} -- and from
conditioning mechanisms such as FiLM~\cite{Perez2018}, originally introduced for visual reasoning. Our
contribution is not the mechanism but its application and a specific insight about \emph{where} it is
necessary: for a comparison-based (argmax or softmax) selector, we show conditioning must modulate the
per-alternative features, because an additive, symmetric signal cancels under the comparison
(\cref{sec:p1design:film}). This connects a general representation-learning tool to a concrete failure mode
in networked decision-making.

\section{Intent-Based Networking}
\label{sec:rel:ibn}

Finally, ``intent'' also names a body of network-management work concerned with translating high-level
operator goals into configuration, surveyed in RFC~9315~\cite{RFC9315}. That community's ``intent'' is an
operator's declarative policy for a whole network; ours is an \emph{application's} per-flow QoS preference
consumed by a learned endpoint controller. The two are complementary rather than competing -- one could
imagine operator intent bounding the space in which application intent is served -- but they operate at
different layers and timescales, and our techniques address the fine-grained, per-flow end of the spectrum.
